cardinalR: Generating interesting high-dimensional data structures

### 2025-07-05 paper-cardinalR.Rmd
Added Trefoil initial description

>>>>

### Trefoil Knot

The **Trefoil Knot** is a closed, nontrivial 1-dimensional manifold embedded in 3D or 4D space. Unlike simple curves like the S-curve or Swiss roll, the trefoil features topological complexity in the form of self-overlaps, making it a valuable test case for evaluating the ability of nonlinear dimensionality reduction (NLDR) methods to preserve global structure, loops, and embeddings in high-dimensional data.

We generate the trefoil structure in **4D space** as the true model, and provide a **stereographically projected 3D version** for visualization purposes. Both versions support the addition of noise dimensions to embed the structure in higher-dimensional ambient spaces, simulating real-world settings.

#### `gen_trefoil4d()`

The function `gen_trefoil4d()` generates a 4-dimensional trefoil knot using two angular parameters, $\theta$ and $\phi$. The knot is constructed on the 3-sphere ($S^3 \subset \mathbb{R}^4$), with a band of thickness around the knot path, controlled by `steps`.

The coordinates are given by:

$$
\begin{aligned}
x_1 &= \cos(\theta) \cdot \cos(\phi), \\
x_2 &= \cos(\theta) \cdot \sin(\phi), \\
x_3 &= \sin(\theta) \cdot \cos(1.5 \cdot \phi), \\
x_4 &= \sin(\theta) \cdot \sin(1.5 \cdot \phi).
\end{aligned}
$$

The number of `theta` and `phi` values is determined by the `steps` and `n` arguments, respectively. Additional independent noise dimensions can be added via `gen_noisedims()` to extend the structure into $p > 4$ dimensions.

```{r}
trefoil4d <- gen_trefoil4d(n = 500, p = 4, steps = 5)
```

The resulting dataset represents a true model $T = (x_1, x_2, x_3, x_4)$, capturing the full 4D knot structure.

#### `gen_trefoil3d()`

The function `gen_trefoil3d()` generates a **3-dimensional stereographic projection** of the 4D trefoil knot, mapping each point from $\mathbb{R}^4$ to $\mathbb{R}^3$ using the transformation:

$$
(x_1', x_2', x_3') = \left( \frac{x_1}{1 - x_4}, \frac{x_2}{1 - x_4}, \frac{x_3}{1 - x_4} \right),
$$

which excludes points where $x_4 = 1$. The projection preserves the knot topology and enables easier visualization and interpretation.

```{r}
trefoil3d <- gen_trefoil3d(n = 500, p = 4, steps = 5)
```

As with the 4D version, this function supports optional noise dimensions to embed the projected knot into higher dimensions.

### 2025-07-05 paper-cardinalR.tex
Set figure option
Updated acknowledgment only with source code
Added Trefoil initial description

\begin{figure}
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-orgcurvybranches-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the `expbranches`, `linearbranches`, `curvybranches`, `orglinearbranches`, `orgcurvybranches` data.}\label{fig:fig-orgcurvybranches-proj}
\end{figure}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-cone-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the `cone` data.}\label{fig:fig-cone-proj}
\end{figure}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-cubehole-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the `gridcube`, `unifcube`, and `cubehole` data.}\label{fig:fig-cubehole-proj}
\end{figure}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-gau-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the `gau` data.}\label{fig:fig-gau-proj}
\end{figure}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-linear-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the `linear` data.}\label{fig:fig-linear-proj}
\end{figure}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-mobius-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the `mobius` data.}\label{fig:fig-mobius-proj}
\end{figure}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-cubic-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the `quadratic` and `cubic` data.}\label{fig:fig-cubic-proj}
\end{figure}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-pyrholes-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the `pyrrect`, `pyrtri`, `pyrstar`, and `pyrholes` data.}\label{fig:fig-pyrholes-proj}
\end{figure}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-scurve-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the `scurve` and `scurvehole` data.}\label{fig:fig-scurve-proj}
\end{figure}

@@

\begin{figure}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-swissroll-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the `swissroll` data.}\label{fig:fig-swissroll-proj}
\end{figure}

@@

@@

\begin{figure}
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-highd-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the `mobiusgau` data. The helical spiral cluster is represented in dark green, the hemisphere cluster in orange, the uniform cube-shaped cluster in purple, the blunted cone cluster in pink, and the Gaussian-shaped cluster in light green.}\label{fig:fig-highd-proj}
\end{figure}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-nldr-layouts-1} \caption{Six different dimension reduction representations of the `mobiusgau` data using default hyperparameter settings: (a) tSNE, (b) UMAP, (c) PAHTE, (d) TriMAP, (e) PaCMAP, and (f) PCA.}\label{fig:fig-nldr-layouts}
\end{figure}

@@

The package is available at \url{https://github.com/JayaniLakshika/cardinalR}, and source material for this paper is available at \url{https://github.com/JayaniLakshika/paper-cardinalR}.

>>>>

\begin{figure}[!]
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-orgcurvybranches-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the `expbranches`, `linearbranches`, `curvybranches`, `orglinearbranches`, `orgcurvybranches` data.}\label{fig:fig-orgcurvybranches-proj}
\end{figure}

@@

\begin{figure}[!]
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-cone-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the `cone` data.}\label{fig:fig-cone-proj}
\end{figure}

@@

\begin{figure}[!]
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-cubehole-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the `gridcube`, `unifcube`, and `cubehole` data.}\label{fig:fig-cubehole-proj}
\end{figure}

@@

\begin{figure}[!]
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-gau-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the `gau` data.}\label{fig:fig-gau-proj}
\end{figure}

@@

\begin{figure}[!]
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-linear-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the `linear` data.}\label{fig:fig-linear-proj}
\end{figure}

@@

\begin{figure}[!]
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-mobius-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the `mobius` data.}\label{fig:fig-mobius-proj}
\end{figure}

@@

\begin{figure}[!]
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-cubic-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the `quadratic` and `cubic` data.}\label{fig:fig-cubic-proj}
\end{figure}

@@

\begin{figure}[!]
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-pyrholes-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the `pyrrect`, `pyrtri`, `pyrstar`, and `pyrholes` data.}\label{fig:fig-pyrholes-proj}
\end{figure}

@@

\begin{figure}[!]
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-scurve-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the `scurve` and `scurvehole` data.}\label{fig:fig-scurve-proj}
\end{figure}

@@

\begin{figure}[!]

@@

\begin{figure}[!]
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-swissroll-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the `swissroll` data.}\label{fig:fig-swissroll-proj}
\end{figure}

@@

\subsubsection{Trefoil Knot}\label{trefoil-knot}

The \textbf{Trefoil Knot} is a closed, nontrivial 1-dimensional manifold embedded in 3D or 4D space. Unlike simple curves like the S-curve or Swiss roll, the trefoil features topological complexity in the form of self-overlaps, making it a valuable test case for evaluating the ability of nonlinear dimensionality reduction (NLDR) methods to preserve global structure, loops, and embeddings in high-dimensional data.

We generate the trefoil structure in \textbf{4D space} as the true model, and provide a \textbf{stereographically projected 3D version} for visualization purposes. Both versions support the addition of noise dimensions to embed the structure in higher-dimensional ambient spaces, simulating real-world settings.

\paragraph{\texorpdfstring{\texttt{gen\_trefoil4d()}}{gen\_trefoil4d()}}\label{gen_trefoil4d}

The function \texttt{gen\_trefoil4d()} generates a 4-dimensional trefoil knot using two angular parameters, \(\theta\) and \(\phi\). The knot is constructed on the 3-sphere (\(S^3 \subset \mathbb{R}^4\)), with a band of thickness around the knot path, controlled by \texttt{steps}.

The coordinates are given by:

\[
\begin{aligned}
x_1 &= \cos(\theta) \cdot \cos(\phi), \\
x_2 &= \cos(\theta) \cdot \sin(\phi), \\
x_3 &= \sin(\theta) \cdot \cos(1.5 \cdot \phi), \\
x_4 &= \sin(\theta) \cdot \sin(1.5 \cdot \phi).
\end{aligned}
\]

The number of \texttt{theta} and \texttt{phi} values is determined by the \texttt{steps} and \texttt{n} arguments, respectively. Additional independent noise dimensions can be added via \texttt{gen\_noisedims()} to extend the structure into \(p > 4\) dimensions.

The resulting dataset represents a true model \(T = (x_1, x_2, x_3, x_4)\), capturing the full 4D knot structure.

\paragraph{\texorpdfstring{\texttt{gen\_trefoil3d()}}{gen\_trefoil3d()}}\label{gen_trefoil3d}

The function \texttt{gen\_trefoil3d()} generates a \textbf{3-dimensional stereographic projection} of the 4D trefoil knot, mapping each point from \(\mathbb{R}^4\) to \(\mathbb{R}^3\) using the transformation:

\[
(x_1', x_2', x_3') = \left( \frac{x_1}{1 - x_4}, \frac{x_2}{1 - x_4}, \frac{x_3}{1 - x_4} \right),
\]

which excludes points where \(x_4 = 1\). The projection preserves the knot topology and enables easier visualization and interpretation.

As with the 4D version, this function supports optional noise dimensions to embed the projected knot into higher dimensions.

@@

\begin{figure}[!]
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-highd-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the `mobiusgau` data. The helical spiral cluster is represented in dark green, the hemisphere cluster in orange, the uniform cube-shaped cluster in purple, the blunted cone cluster in pink, and the Gaussian-shaped cluster in light green.}\label{fig:fig-highd-proj}
\end{figure}

@@

\begin{figure}[!]
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-nldr-layouts-1} \caption{Six different dimension reduction representations of the `mobiusgau` data using default hyperparameter settings: (a) tSNE, (b) UMAP, (c) PAHTE, (d) TriMAP, (e) PaCMAP, and (f) PCA.}\label{fig:fig-nldr-layouts}
\end{figure}

@@

The source material for this paper is available at \url{https://github.com/JayaniLakshika/paper-cardinalR}.
